Weighted Model Counting Without Parameter Variables

### 2021-02-20 doc/paper3/paper.tex
The big proof is mostly complete!

\paragraph{Notation.} For any propositional formula $\phi$ and $p, q \in
\mathbb{R}$, let $[\phi]^p_q\colon 2^X \to \mathbb{R}$ be a
pseudo-Boolean function defined as
\[
  [\phi]^p_q(Y) \coloneqq
  \begin{cases}
    p & \text{if } Y \models \phi \\
    q & \text{otherwise}
  \end{cases}
\]
for any $Y \subseteq X$.

% NOTE: weights on literals other than the positive literals that correspond to
% variables in $X_P$ are redundant as they either are equal to one or duplicate
% an already-defined weight.
\begin{definition}[old WMC instance]
  An \emph{old WMC instance} is a tuple $(\phi, X_I, X_P, w)$, where $X_I$ is
  the set of indicator variables, $X_P$ is the set of parameter variables (with
  $X_I \cap X_P = \emptyset$), $\phi$ is a propositional formula in CNF over
  $X_I \cup X_P$, and $w\colon X_I \cup X_P \cup \{\neg x \mid x \in X_I \cup
  X_P\} \to \mathbb{R}$ is the weight function. The \emph{answer} of the
  instance is $\sum_{Y \models \phi} \prod_{Y \models l} w(l)$.
\end{definition} % TODO: add a citation for this.

@@

\paragraph{Notes.}
\begin{itemize}
\item Apparently, the \textsc{DPMC} paper already shows that taking the first
  offered decomposition tree is best.
\item It is already well-known that WMC is FPT.
\end{itemize}

@@

\begin{theorem}[Early Projection
  \cite{DBLP:conf/aaai/DudekPV20,DBLP:conf/cp/DudekPV20},
  verbatim] \label{thm:early}
  Let $X$ and $Y$ be sets of variables. For all functions $f\colon 2^X \to
  \mathbb{R}$ and $g\colon 2^Y \to \mathbb{R}$, if $x \in X \setminus Y$, then
  $\exists_x (f \cdot g) = (\exists_x f) \cdot g$.
\end{theorem}
% TODO: mention that formulas, clauses, and models are all treated as sets

\begin{lemma} \label{lemma:sum}
  For any pseudo-Boolean function $f\colon 2^X \to \mathbb{R}$, $(\exists_X
  f)(\emptyset) = \sum_{Y \subseteq X} f(Y)$.
\end{lemma}
\begin{proof}
  For base case, let $X = \{x\}$. Then
  \[
    \exists_xf = f|_{x=1} + f|_{x=0},
  \]
  by \cref{def:projection} and so
  \[
    (\exists_xf)(\emptyset) = f|_{x=1}(\emptyset) + f|_{x=0}(\emptyset)
  \]
  by \cref{def:operations}. Then
  \[
    f|_{x=1}(\emptyset) = f(\{x\}), \quad \text{and} \quad f|_{x=0}(\emptyset) =
    f(\emptyset)
  \]
  by \cref{def:restriction}. It follows that
  \[
    (\exists_xf)(\emptyset) = f(\{x\}) + f(\emptyset) = \sum_{Y \subseteq \{x\}}
    f(Y)
  \]
  as required. This easily extends to $|X| > 1$ by the definition of projection
  on sets of variables. % TODO: once I've defined it, add a reference here.
\end{proof}
% TODO: must mention that operations like +/x are commutative, associative, etc.

@@

\begin{theorem}[Correctness]
  \Cref{alg:transformation}, when given an old WMC instance $(\phi, X_I, X_P,
  w)$, returns a new WMC instance with the same answer, provided the following
  conditions are satisfied:
  \begin{enumerate}
  \item for all indicator variables $i \in X_I$, $w(i) = w(\neg i) = 1$,
  \item and either
    \begin{enumerate}
    \item for all parameter variables $p \in X_P$, there is a non-empty family
      of literals $(l_i)_{i=1}^n$ such that \label{condition:d02}
      \begin{enumerate}
      \item $w(\neg p) = 1$,
      \item $l_i \in X_I$ or $\neg l_i \in X_I$ for all $i = 1, \dots, n$,
        \label{condition:equivalence1}
      \item and $\{ c \in \phi \mid p \in c \text{ or } \neg p \in c \} =
        \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
        \mid i = 1, \dots, n \}$; \label{condition:equivalence2}
      \end{enumerate}
    \item or for all parameter variables $p \in X_P$, \label{condition:bklm16}
      \begin{enumerate}
      \item $w(p) + w(\neg p) = 1$,
      \item for any clause $c \in \phi$, $|c \cap X_P| \le 1$,
      \item there is no clause $c \in \phi$ such that $\neg p \in c$,
      \item and for any pair of clauses $c, d \in \phi$ such that $c \ne d$, $p
        \in c$ and $p \in d$, $\mathtt{Vars}(c) \cap \mathtt{Vars}(d) = \{ p
        \}$.
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}
\end{theorem} % TODO: add parentheses around products
% TODO: maybe give a name to (2) since it's wrong to call it an equation
\begin{proof} % TODO: check if each condition is actually used
  % TODO: start here: consider how the transformation changes the 'normal form'.
  % We will show that every iteration keeps the answer the same as in the normal form.
  By \cref{prop:equivalence},
  \begin{equation}
    \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
          x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
      1\right) \label{eq:new_wmc}
  \end{equation}
  is a new WMC instance with the same answer as the given old WMC instance. By
  \cref{def:new_wmc}, its answer is
  \begin{equation} \label{eq:answer}
    \left(\exists_{X_I \cup X_P} \prod_{c \in \phi} [c]_0^1 \prod_{x \in X_I
        \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset)
  \end{equation}
  Since both \cref{condition:d02,condition:bklm16} ensure that each clause in
  $\phi$ has at most one parameter variable, we can partition $\phi$ into
  $\phi_* \coloneqq \{c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \emptyset \}$
  and $\phi_p \coloneqq \{ c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \{ p \}
  \}$ for all $p \in X_P$. We can then use \cref{thm:early} to reorder
  \cref{eq:answer} into
  \[
    \left(\exists_{X_I} \left( \prod_{x \in X_I} [x]_{w(\neg x)}^{w(x)} \right)
      \left( \prod_{c \in \phi_*} [c]_0^1 \right) \prod_{p \in X_P} \exists_p
      [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 \right)(\emptyset).
  \]

@@

  If $w(p) \ne 1$, then
  \[
    F_p = \left\{ \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} \right\},
  \]
  and so we want to show that
  \[
    \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} = \exists_p [p]_1^{w(p)}
    \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1.
  \]
  By \cref{def:projection},
  \begin{align*}
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 &= \left.\left( [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 \right)\right|_{p=1} + \left.\left([p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 \right)\right|_{p=0} \\
                                                                                      &= w(p) \cdot \left[ \bigwedge_{i=1}^n l_i \right]_0^1 + \left[ \bigwedge_{i=1}^n l_i \right]_1^0 = \left[ \bigwedge_{i=1}^n l_i \right]_1^{w(p)}
  \end{align*}
  as required. This finishes the proof of the correctness of the first `foreach'
  loop under \cref{condition:d02}.

  Now let us assume \cref{condition:bklm16}. We still want to prove
  \cref{eq:to_show}.
\end{proof}
% TODO: now consider the second for loop
% TODO: condition 1 is necessary in both cases because we're ignore the weights
% of indicator variables

@@

\begin{definition} \label{def:operations}
  Let $f, g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions. Operations
  such as addition and multiplication are defined pointwise as
  \[
    (f+g)(Y) \coloneqq f(Y)+g(Y), \quad \text{and} \quad (fg)(Y) \coloneqq
    f(Y)g(Y)
  \]
  for all $Y \subseteq X$.
\end{definition}

>>>>

\paragraph{Notation.} For any propositional formula $\phi$ over a set of
variables $X$ and $p, q \in \mathbb{R}$, let $[\phi]^p_q\colon 2^X \to
\mathbb{R}$ be a pseudo-Boolean function defined as
\[
  [\phi]^p_q(Y) \coloneqq
  \begin{cases}
    p & \text{if } Y \models \phi \\
    q & \text{otherwise}
  \end{cases}
\]
for any $Y \subseteq X$.

% NOTE: ...that we will need.
% NOTE: The proof immediately comes from the definition (of operations).
\begin{proposition}[Basic properties] \label{prop:basic}
  For any propositional formulas $\phi$ and $\psi$, and $a, b, c, d \in
  \mathbb{R}$,
  \begin{itemize}
  \item $[\phi]^a_b = [\neg \phi]^b_a$;
  \item $c + [\phi]^a_b = [\phi]^{a+c}_{b+c}$;
  \item $c \cdot [\phi]^a_b = [\phi]^{ac}_{bc}$;
  \item $[\phi]^a_b \cdot [\phi]^c_d = [\phi]^{ac}_{bd}$;
  \item $[\phi]^1_0 \cdot [\psi]_0^1 = [\phi \land \psi]_0^1$.
  \end{itemize}
\end{proposition}

\begin{definition}[old WMC instance]
  An \emph{old WMC instance} is a tuple $(\phi, X_I, X_P, w)$, where $X_I$ is
  the set of indicator variables, $X_P$ is the set of parameter variables (with
  $X_I \cap X_P = \emptyset$), $\phi$ is a propositional formula in CNF over
  $X_I \cup X_P$, and $w\colon X_I \cup X_P \cup \{\neg x \mid x \in X_I \cup
  X_P\} \to \mathbb{R}$ is the weight function. The \emph{answer} of the
  instance is $\sum_{Y \models \phi} \prod_{Y \models l} w(l)$.
\end{definition} % TODO: add a citation for this.

@@

\paragraph{Notes.}
\begin{itemize}
\item Apparently, the \textsc{DPMC} paper already shows that taking the first
  offered decomposition tree is best.
\item It is already well-known that WMC is FPT.
\item Weights on literals other than the positive literals that correspond to
  variables in $X_P$ are redundant as they either are equal to one or duplicate
  an already-defined weight.
\item Mention that formulas, clauses, and models are all treated as sets.
\end{itemize}

@@

\begin{theorem}[Early Projection
  \cite{DBLP:conf/aaai/DudekPV20,DBLP:conf/cp/DudekPV20},
  verbatim] \label{thm:early}
  Let $X$ and $Y$ be sets of variables. For all functions $f\colon 2^X \to
  \mathbb{R}$ and $g\colon 2^Y \to \mathbb{R}$, if $x \in X \setminus Y$, then
  $\exists_x (f \cdot g) = (\exists_x f) \cdot g$.
\end{theorem}

\begin{lemma} \label{lemma:sum}
  For any pseudo-Boolean function $f\colon 2^X \to \mathbb{R}$, $(\exists_X
  f)(\emptyset) = \sum_{Y \subseteq X} f(Y)$.
\end{lemma}
\begin{proof}
  For base case, let $X = \{x\}$. Then
  \[
    \exists_xf = f|_{x=1} + f|_{x=0},
  \]
  by \cref{def:projection} and so
  \[
    (\exists_xf)(\emptyset) = f|_{x=1}(\emptyset) + f|_{x=0}(\emptyset)
  \]
  by \cref{def:operations}. Then
  \[
    f|_{x=1}(\emptyset) = f(\{x\}), \quad \text{and} \quad f|_{x=0}(\emptyset) =
    f(\emptyset)
  \]
  by \cref{def:restriction}. It follows that
  \[
    (\exists_xf)(\emptyset) = f(\{x\}) + f(\emptyset) = \sum_{Y \subseteq \{x\}}
    f(Y)
  \]
  as required. This easily extends to $|X| > 1$ by the definition of projection
  on sets of variables. % TODO: once I've defined it, add a reference here.
\end{proof}

@@

\begin{theorem}[Correctness]
  \Cref{alg:transformation}, when given an old WMC instance $(\phi, X_I, X_P,
  w)$, returns a new WMC instance with the same answer, provided the following
  conditions are satisfied:
  \begin{enumerate}
  \item for all indicator variables $i \in X_I$, $w(i) = w(\neg i) = 1$,
  \item and either
    \begin{enumerate}
    \item for all parameter variables $p \in X_P$, there is a non-empty family
      of literals $(l_i)_{i=1}^n$ such that \label{condition:d02}
      \begin{enumerate}
      \item $w(\neg p) = 1$,
      \item $l_i \in X_I$ or $\neg l_i \in X_I$ for all $i = 1, \dots, n$,
        \label{condition:equivalence1}
      \item and $\{ c \in \phi \mid p \in c \text{ or } \neg p \in c \} =
        \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
        \mid i = 1, \dots, n \}$; \label{condition:equivalence2}
      \end{enumerate}
    \item or for all parameter variables $p \in X_P$, \label{condition:bklm16}
      \begin{enumerate}
      \item $w(p) + w(\neg p) = 1$,
      \item for any clause $c \in \phi$, $|c \cap X_P| \le 1$,
      \item there is no clause $c \in \phi$ such that $\neg p \in c$,
      \item if $\{ p \} \in \phi$, then there is no clause $c \in \phi$ such
        that $c \ne \{ p \}$ and $p \in c$, \label{cond:just_parameter}
        % 'prime implicants are logically disjoint'
      \item and for any $c, d \in \phi$ such that $c \ne d$, $p \in c$ and $p
        \in d$, $\bigwedge_{l \in c \setminus \{ p \}} \neg l \land \bigwedge_{l
          \in d \setminus \{ p \}} \neg l$ is false. \label{cond:disjoint}
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}
\end{theorem} % TODO: add parentheses around products
% TODO: maybe give a name to (2) since it's wrong to call it an equation
% TODO: check if each condition is actually used. Maybe turn this into a
% paragraph that gives an overview of the proof.
\begin{proof}
  By \cref{prop:equivalence},
  \begin{equation}
    \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
          x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
      1\right) \label{eq:new_wmc}
  \end{equation}
  is a new WMC instance with the same answer as the given old WMC instance. By
  \cref{def:new_wmc}, its answer is
  \begin{equation} \label{eq:answer}
    \left(\exists_{X_I \cup X_P} \prod_{c \in \phi} [c]_0^1 \prod_{x \in X_I
        \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset)
  \end{equation}
  Since both \cref{condition:d02,condition:bklm16} ensure that each clause in
  $\phi$ has at most one parameter variable, we can partition $\phi$ into
  $\phi_* \coloneqq \{c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \emptyset \}$
  and $\phi_p \coloneqq \{ c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \{ p \}
  \}$ for all $p \in X_P$. We can then use \cref{thm:early} to reorder
  \cref{eq:answer} into
  \[
    \left(\exists_{X_I} \left( \prod_{x \in X_I} [x]_{w(\neg x)}^{w(x)} \right)
      \left( \prod_{c \in \phi_*} [c]_0^1 \right) \prod_{p \in X_P} \exists_p
      [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 \right)(\emptyset).
  \]

@@

  If $w(p) \ne 1$, then
  \[
    F_p = \left\{ \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} \right\},
  \]
  and $\omega = 1$, and so we want to show that
  \[
    \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} = \exists_p [p]_1^{w(p)}
    \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1.
  \]
  By \cref{def:projection},
  \begin{align*}
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 &= \left.\left( [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 \right)\right|_{p=1} + \left.\left([p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 \right)\right|_{p=0} \\
                                                                                      &= w(p) \cdot \left[ \bigwedge_{i=1}^n l_i \right]_0^1 + \left[ \bigwedge_{i=1}^n l_i \right]_1^0 = \left[ \bigwedge_{i=1}^n l_i \right]_1^{w(p)}
  \end{align*}
  as required. This finishes the proof of the correctness of the first `foreach'
  loop under \cref{condition:d02}.

  Now let us assume \cref{condition:bklm16}. We still want to prove
  \cref{eq:to_show}. If $w(p) = 1$, then $F_p = \emptyset$, and $\omega = 1$,
  and so the left-hand side of \cref{eq:to_show} is equal to one. Then the
  right-hand side is
  \[
    \exists_p [p]_0^1 \prod_{c \in \phi_p} [c]_0^1 = \exists_p \left[ p \land
      \bigwedge_{c \in \phi_p} c \right]_0^1 = \exists_p [p]_0^1 = 0 + 1 = 1
  \]
  by \cref{prop:basic,def:projection}, and because $p \in c$ for every clause $c
  \in \phi_p$.

  If $w(p) \ne 1$, and $\{ p \} \in \phi_p$, then, by
  \cref{cond:just_parameter}, $\phi_p = \{ \{ p \} \}$, and
  \cref{alg:transformation} produces $F_p = \emptyset$ and $\omega = w(p)$, and
  so
  \[
    \omega \prod_{f \in F_p} f = w(p).
  \]
  Whereas on the other side of \cref{eq:to_show},
  \[
    \exists_p [p]_{w(\neg p)}^{w(p)} [p]_0^1 = \exists_p [p]^{w(p)}_0 = w(p)
  \]
  by \cref{prop:basic,def:projection}.

  The only remaining case is when $w(p) \ne 1$ and $\{ p \} \not \in \phi_p$.
  Then $\omega = 1$, and
  \[
    F_p = \left\{ \left[\bigwedge_{l \in c \setminus \{ p \}} \neg
        l\right]_1^{w(p)} \;\middle|\; c \in \phi_p \right\},
  \]
  so need to show that
  \[
    \prod_{c \in \phi_p} \left[\bigwedge_{l \in c \setminus \{ p \}} \neg
      l\right]_1^{w(p)} = \exists_p [p]_{1-w(p)}^{w(p)} \prod_{c \in \phi_p}
    [c]_0^1.
  \]
  We can rearrange the right-hand-side as
  \begin{align*}
    \exists_p [p]_{1-w(p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 &= \exists_p [p]_{1-w(p)}^{w(p)} \left[\bigwedge_{c \in \phi_p} c\right]_0^1 = \exists_p [p]_{1-w(p)}^{w(p)} \left[ p \lor \bigwedge_{c \in \phi_p} c \setminus \{ p \} \right]_0^1 \\
                                                               &= w(p) + (1-w(p)) \left[ \bigwedge_{c \in \phi_p} c \setminus \{ p \} \right]_0^1 = w(p) + \left[ \bigwedge_{c \in \phi_p} c \setminus \{ p \} \right]_0^{1-w(p)} \\
                                                               &= \left[ \bigwedge_{c \in \phi_p} c \setminus \{ p \} \right]_{w(p)}^1 = \left[ \neg \bigwedge_{c \in \phi_p} c \setminus \{ p \} \right]_1^{w(p)} = \left[ \bigvee_{c \in \phi_p} \neg(c \setminus \{ p \}) \right]_1^{w(p)} \\
                                                               &= \left[ \bigvee_{c \in \phi_p} \neg \bigvee_{l \in c \setminus \{ p \}} l \right]_1^{w(p)} = \left[ \bigvee_{c \in \phi_p} \bigwedge_{l \in c \setminus \{ p \}} \neg l \right]_1^{w(p)}
  \end{align*}
   by \cref{prop:basic,def:projection}. By \cref{cond:disjoint}, $\bigwedge_{l
     \in c \setminus \{ p \}} \neg l$ can be true for at most one $c \in
   \phi_p$, and so
   \[
     \left[ \bigvee_{c \in \phi_p} \bigwedge_{l \in c \setminus \{ p \}} \neg l
     \right]_1^{w(p)} = \prod_{c \in \phi_p} \left[ \bigwedge_{l \in c \setminus
         \{ p \}} \neg l \right]_1^{w(p)}
   \]
   which is exactly what we needed to show.
\end{proof}
% TODO: now consider the second for loop
% TODO: condition 1 is necessary in both cases because we're ignore the weights
% of indicator variables
% NOTE: benefits of having the proof in the paper:
% 1) it puts all the encodings on a common ground
% 2) it illustrates the convenience of our notation for reasoning about (certain
% types of) pseudo-Boolean functions (stress this!!!)
% 3) it's too big and too important to be left for the appendix
% NOTE: we don't make a distinction between a real number and a (PB) function
% that always returns that number.
% NOTE: projection is assumed to have lowest precedence (i.e., \exists_x fg =
% \exists_x (fg) and not (\exists_x f)g).
% TODO: need another basic property: (fg)|_{x=i} = f|_{x=i} * g|_{x=i} for i=0, 1.
% TODO: \times -> \cdot

@@

\begin{definition} \label{def:operations}
  Let $f, g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions. Operations
  such as addition and multiplication are defined pointwise as
  \[
    (f+g)(Y) \coloneqq f(Y)+g(Y), \quad \text{and} \quad (fg)(Y) \coloneqq
    f(Y)g(Y)
  \]
  for all $Y \subseteq X$. This means that binary operations on pseudo-Boolean
  functions inherit properties such as associativity and commutativity.
\end{definition}
